---
output:
  github_document:
    pandoc_args: ["--markdown-headings=atx"]
vignette: >
  %\VignetteIndexEntry{1. Fetching data directly from a JSON spec}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# 1. Fetching data directly from a JSON spec

## Overview

The simplest way to get data from Stat-Xplore is to pass an existing JSON
spec file directly to `fetch_table()`. This vignette shows that workflow
end-to-end using the Attendance Allowance (ACC) dataset.

The JSON spec used here selects:

- **Database**: Attendance Allowance (ACC)
- **Measure**: ACC claimant count (`V_F_ACC`)
- **Age**: single year of age (16–65 + 99)
- **Employment status**: employed, unemployed, unknown
- **Geography**: England and Wales (country level)
- **Date**: August 2013, August 2019, August 2022

```json
{
  "database": "str:database:ACC",
  "measures": ["str:count:ACC:V_F_ACC"],
  "recodes": {
    "str:field:ACC:V_F_ACC:AGE": {
      "map": [["str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:16"], ...],
      "total": true
    },
    ...
  },
  "dimensions": [
    ["str:field:ACC:V_F_ACC:AGE"],
    ["str:field:ACC:F_ACC_DATE_new:DATE_NAME"],
    ["str:field:ACC:V_F_ACC:WARD_CODE"],
    ["str:field:ACC:V_F_ACC:EMP"]
  ]
}
```

## Step 1 – Load the package and set your API key

```{r load-pkg}
library(statxplorerplus)

# Load your API key from a text file (one line, no trailing whitespace)
load_api_key("path/to/apikey.txt")
```

## Step 2 – Fetch data from the JSON file

Pass the path to your saved JSON spec to `fetch_table()`. The function
sends the query to the Stat-Xplore API and returns a tidy data frame by
default.

```{r fetch}
json_path <- system.file("extdata", "acc_ew_by_single_year_of_age.json",
                         package = "statxplorerplus")
acc_data <- fetch_table(filename = json_path)

head(acc_data)
```

```
#>   V_F_ACC    AGE DATE_NAME   WARD_CODE EMP         value
#>   <chr>    <chr> <chr>       <chr>     <chr>       <dbl>
#> 1 ACC count 16  August 2013 England   Employed       12
#> 2 ACC count 16  August 2013 England   Unemployed      3
#> ...
```

## Step 3 – Inspect the full results object (optional)

If you need access to the raw measures, field labels, or URIs returned by
the API, set `only_keep_data = FALSE`:

```{r fetch-full}
acc_results <- fetch_table(
  filename       = system.file("extdata", "acc_ew_by_single_year_of_age.json",
                               package = "statxplorerplus"),
  only_keep_data = FALSE
)

# Available elements
names(acc_results)
#> [1] "measures" "fields"   "items"    "uris"     "dfs"

# Human-readable field labels
acc_results$fields
```

## Step 4 – Drop totals (optional)

Stat-Xplore can return a "Total" row for each recode dimension. Pass
`drop_total_rows = TRUE` to remove them automatically:

```{r fetch-no-totals}
acc_data_clean <- fetch_table(
  filename        = system.file("extdata", "acc_ew_by_single_year_of_age.json",
                                package = "statxplorerplus"),
  drop_total_rows = TRUE
)
```

## Summary

| Function | Purpose |
|---|---|
| `load_api_key()` | Load API key from file |
| `fetch_table(filename = ...)` | Query Stat-Xplore and return a data frame |
| `fetch_table(..., only_keep_data = FALSE)` | Return the full results list |
| `fetch_table(..., drop_total_rows = TRUE)` | Remove "Total" rows from results |
