---
output:
  github_document:
    pandoc_args: ["--markdown-headings=atx"]
vignette: >
  %\VignetteIndexEntry{2. Subsetting API calls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# 2. Subsetting API calls

## Overview

A JSON spec exported from Stat-Xplore may contain far more values than you
need for a particular analysis. Rather than sending the full query to the
API, you can convert the JSON to a spec table, filter it down to only the
rows you want, and then fetch just that subset of the data.

This vignette shows how to load a JSON spec, isolate a single age value
(age 16), and download only that slice of the data.

## Step 1 – Load the package and set your API key

```{r load-pkg}
library(statxplorerplus)
library(dplyr)

load_api_key("path/to/apikey.txt")
```

## Step 2 – Convert the JSON to a spec table and add labels

```{r convert}
json_path <- "path/to/acc_ew_by_single_year_of_age_pre19.json"

spec_tbl <- convert_json_to_spec_table(json_path) |>
  add_labels_and_locations_to_spec_table()
```

The full spec table contains rows for all 51 single years of age as well as
rows for the geography, employment, and date fields. You can inspect which
fields and values are included:

```{r inspect}
spec_tbl |>
  count(field_label)
```

```
#> # A tibble: 4 × 2
#>   field_label                        n
#>   <chr>                          <int>
#> 1 Age of Claimant (Single Years)    51
#> 2 Employment Indicator               3
#> 3 Month                              3
#> 4 Region/Country                     2
```

## Step 3 – Filter to age 16 only

Keep only the row for age 16 from the age field, leaving all other fields
unchanged:

```{r subset}
spec_tbl_age16 <- spec_tbl |>
  filter(
    field_label != "Age of Claimant (Single Years)" |
      value_code == "16"
  )

spec_tbl_age16 |>
  select(field_label, value_label, value_code)
```

```
#> # A tibble: 9 × 3
#>   field_label                    value_label  value_code
#>   <chr>                          <chr>        <chr>
#> 1 Age of Claimant (Single Years) 16           16
#> 2 Employment Indicator           Employed     0
#> 3 Employment Indicator           Unemployed   1
#> 4 Employment Indicator           Unknown      99
#> 5 Month                          August 2013  201308
#> 6 Month                          August 2019  201908
#> 7 Month                          August 2022  202208
#> 8 Region/Country                 England      E92000001
#> 9 Region/Country                 Wales        W92000004
```

The spec table now has just one age row instead of 51, which will make the
API call much faster.

## Step 4 – Fetch the data

```{r fetch}
acc_age16 <- fetch_data_from_spec_table(spec_tbl_age16)

acc_age16
```

```
#>   V_F_ACC   AGE DATE_NAME    WARD_CODE EMP        value
#>   <chr>     <chr> <chr>      <chr>     <chr>      <dbl>
#> 1 ACC count 16   August 2013 England   Employed      12
#> 2 ACC count 16   August 2013 England   Unemployed     3
#> 3 ACC count 16   August 2013 Wales     Employed       2
#> ...
```

## Summary

| Step | Code | Purpose |
|---|---|---|
| 1 | `convert_json_to_spec_table()` | Parse JSON → spec table |
| 2 | `add_labels_and_locations_to_spec_table()` | Add human-readable labels |
| 3 | `filter()` | Keep only the values you need |
| 4 | `fetch_data_from_spec_table()` | Fetch the subset from the API |
