---
output:
  github_document:
    pandoc_args: ["--markdown-headings=atx"]
vignette: >
  %\VignetteIndexEntry{3. Modifying an existing query}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# 3. Modifying an existing query

## Overview

Once a query is loaded into a spec table, you can change any of its values
using standard `dplyr` operations — no need to edit the JSON file directly.

This vignette starts with a query that downloads Attendance Allowance (ACC)
claimant counts for **16-year-olds in England** and modifies it to instead
download the same data for **17-year-olds in Wales**.

## Step 1 – Load the package and set your API key

```{r load-pkg}
library(statxplorerplus)
library(dplyr)

load_api_key("path/to/apikey.txt")
```

## Step 2 – Load the original query

Convert the JSON to a spec table and add labels so you can see what each
row refers to:

```{r convert}
json_path <- "path/to/acc_ew_by_single_year_of_age_pre19.json"

spec_tbl <- convert_json_to_spec_table(json_path) |>
  add_labels_and_locations_to_spec_table()
```

Check the current age and geography selections:

```{r inspect}
spec_tbl |>
  filter(field_label %in% c("Age of Claimant (Single Years)",
                             "Region/Country")) |>
  select(field_label, value_label, value_code)
```

```
#> # A tibble: 53 × 3
#>   field_label                    value_label value_code
#>   <chr>                          <chr>       <chr>
#> 1 Age of Claimant (Single Years) 16          16
#> 2 Age of Claimant (Single Years) 17          17
#> ...
#> 52 Region/Country                 England     E92000001
#> 53 Region/Country                 Wales       W92000004
```

## Step 3 – Modify the query

Replace age 16 with age 17 in the age field, and replace England with Wales
in the geography field:

```{r modify}
spec_tbl_modified <- spec_tbl |>
  filter(
    # Keep only age 17 (drop all other ages)
    field_label != "Age of Claimant (Single Years)" |
      value_code == "17",
    # Keep only Wales (drop England)
    field_label != "Region/Country" |
      value_code == "W92000004"
  )

spec_tbl_modified |>
  filter(field_label %in% c("Age of Claimant (Single Years)",
                             "Region/Country")) |>
  select(field_label, value_label, value_code)
```

```
#> # A tibble: 2 × 3
#>   field_label                    value_label value_code
#>   <chr>                          <chr>       <chr>
#> 1 Age of Claimant (Single Years) 17          17
#> 2 Region/Country                 Wales       W92000004
```

## Step 4 – Fetch the modified data

```{r fetch}
acc_17_wales <- fetch_data_from_spec_table(spec_tbl_modified)

acc_17_wales
```

```
#>   V_F_ACC   AGE DATE_NAME    WARD_CODE EMP        value
#>   <chr>     <chr> <chr>      <chr>     <chr>      <dbl>
#> 1 ACC count 17   August 2013 Wales     Employed       2
#> 2 ACC count 17   August 2013 Wales     Unemployed     1
#> 3 ACC count 17   August 2019 Wales     Employed       3
#> ...
```

## Step 5 – (Optional) export the modified query as a new JSON

Save the modified spec table as a JSON file so it can be reused later:

```{r export}
modified_list <- convert_spec_table_to_list(spec_tbl_modified)

export_json(
  .json        = modified_list,
  .output_path = "path/to/acc_17_wales.json"
)
```

## Summary

| Step | Code | Purpose |
|---|---|---|
| 1 | `convert_json_to_spec_table()` | Parse JSON → spec table |
| 2 | `add_labels_and_locations_to_spec_table()` | Add labels to identify rows |
| 3 | `filter()` | Swap out the age and geography values |
| 4 | `fetch_data_from_spec_table()` | Fetch the modified query |
| 5 | `convert_spec_table_to_list()` + `export_json()` | Save as a new JSON |
