---
output:
  github_document:
    pandoc_args: ["--markdown-headings=atx"]
vignette: >
  %\VignetteIndexEntry{4. Grouping variables in a spec table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# 4. Grouping variables in a spec table

## Why group variables?

Stat-Xplore provides data broken down by individual values — for example,
single years of age (16, 17, 18, …). Sometimes you want to combine these into
broader categories that are not available as built-in groupings in Stat-Xplore
itself. For instance, you might want age bands such as 16–34, 35–54, and 55+
rather than individual years.

`statxplorerplus` supports this through the `value_group` column of a **spec
table**: rows that share the same `value_group` string are sent to the API as a
single grouped category.

### What does grouping do to the returned values?

The effect depends on what you are asking the API for:

- **Counts** — the API returns the *sum* of the individual values within the
  group. So if age 16 has 200 claimants and age 17 has 180, a "16–17" group
  would return 380.
- **Means / averages** — the API returns the *weighted average* across the
  individual values within the group, weighted by the counts of each individual
  value.

### Overview

This vignette shows how to:

1. Convert a JSON spec to a spec table.
2. Add human-readable labels so you can see what each row refers to.
3. Assign `value_group` labels to create age-band groupings.
4. Fetch the grouped data.

## Step 1 – Load the package and set your API key

```{r load-pkg}
library(statxplorerplus)
library(dplyr)

load_api_key("path/to/apikey.txt")
```

## Step 2 – Convert the JSON to a spec table

```{r convert}
json_path <- system.file("extdata", "acc_ew_by_single_year_of_age.json",
                         package = "statxplorerplus")

spec_tbl <- convert_json_to_spec_table(json_path)

spec_tbl
```

```
#> # A tibble: 59 × 6
#>   value_id                                      field_id                  value_group database_id      measure_id            valueset_id
#>   <chr>                                         <chr>                     <chr>       <chr>            <chr>                 <chr>
#> 1 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:16 str:field:ACC:V_F_ACC:AGE <NA>        str:database:ACC str:count:ACC:V_F_ACC str:valueset:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE
#> 2 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:17 str:field:ACC:V_F_ACC:AGE <NA>        str:database:ACC str:count:ACC:V_F_ACC str:valueset:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE
#> ...
```

## Step 3 – Add labels so you can identify the values

```{r add-labels}
spec_tbl_labelled <- spec_tbl |>
  add_labels_and_locations_to_spec_table()

spec_tbl_labelled |>
  select(field_label, value_label, value_code)
```

```
#> # A tibble: 59 × 3
#>   field_label                 value_label value_code
#>   <chr>                       <chr>       <chr>
#> 1 Age (bands and single year) 16          16
#> 2 Age (bands and single year) 17          17
#> 3 Age (bands and single year) 18          18
#> ...
```

## Step 4 – Assign age-band groupings via `value_group`

Add a `value_group` column to define how individual values are collapsed.
Rows without a `value_group` (i.e. `NA`) are kept as individual categories.
Here we group single years of age into three broad bands:

```{r group-age}
spec_tbl_grouped <- spec_tbl_labelled |>
  mutate(
    value_group = case_when(
      field_label == "Age (bands and single year)" &
        as.integer(value_code) %in% 16:34 ~ "16-34",
      field_label == "Age (bands and single year)" &
        as.integer(value_code) %in% 35:54 ~ "35-54",
      field_label == "Age (bands and single year)" &
        as.integer(value_code) >= 55      ~ "55+",
      TRUE ~ NA_character_
    )
  )

spec_tbl_grouped |>
  filter(field_label == "Age (bands and single year)") |>
  select(value_label, value_code, value_group)
```

```
#> # A tibble: 51 × 3
#>   value_label value_code value_group
#>   <chr>       <chr>      <chr>
#> 1 16          16         16-34
#> 2 17          17         16-34
#> ...
#> 20 35          35         35-54
#> ...
#> 40 55          55         55+
```

## Step 5 – Fetch the grouped data

Pass the spec table (with `value_group` populated) directly to
`fetch_data_from_spec_table()`. The groupings are applied automatically
before the query is sent.

```{r fetch}
acc_grouped <- fetch_data_from_spec_table(spec_tbl_grouped)

head(acc_grouped)
```

```
#> # A tibble: 6 × 5
#>   `Age (bands and single year)` Month       `National - Regional - LA - Wards` `Employment Indicator` `Alternative Claimant Count`
#>   <chr>                         <chr>       <chr>                              <chr>                                         <dbl>
#> 1 16-34                         August 2013 England                            Not in employment                            704890
#> 2 16-34                         August 2013 England                            In employment                                 82665
#> 3 16-34                         August 2013 England                            Not available                                     0
#> 4 16-34                         August 2013 Wales                              Not in employment                             46972
#> 5 16-34                         August 2013 Wales                              In employment                                  5060
#> 6 16-34                         August 2013 Wales                              Not available                                     0
```

## How grouping works under the hood

When `value_group` is set, `fetch_data_from_spec_table()` calls
`construct_custom_var_mapping_from_spec_table()` to build a named list
that maps group labels to the individual Stat-Xplore value IDs. This list
is forwarded to `extract_results()` so that the API response is correctly
labelled with your custom group names instead of the raw IDs.

## Summary

| Step | Function | Purpose |
|---|---|---|
| 1 | `convert_json_to_spec_table()` | Parse JSON → spec table |
| 2 | `add_labels_and_locations_to_spec_table()` | Add human-readable labels |
| 3 | `mutate(value_group = ...)` | Define age-band groupings |
| 4 | `fetch_data_from_spec_table()` | Fetch grouped data from API |
