---
output:
  github_document:
    pandoc_args: ["--markdown-headings=atx"]
vignette: >
  %\VignetteIndexEntry{5. Updating a spec to the latest available data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# 5. Updating a spec to the latest available data

## Overview

Stat-Xplore data is updated regularly. A JSON spec you downloaded months ago
will only contain the dates that were available at that time. Rather than
manually editing the JSON every time new data is released,
`update_spec_table_to_latest_data()` queries the Stat-Xplore schema and
appends date values automatically.

This vignette shows the full workflow:

1. Convert a JSON to a spec table.
2. Inspect which dates are currently in the spec.
3. Remove the most recent date to simulate a spec that is not fully up to date.
4. Update the spec to include all dates newer than those already in the spec.
5. Fetch the updated data and check which months are covered.

## Step 1 – Load the package and set your API key

```{r load-pkg}
library(statxplorerplus)
library(dplyr)

load_api_key("path/to/apikey.txt")
```

## Step 2 – Convert the JSON to a spec table and add labels

```{r convert}
json_path <- system.file("extdata", "acc_ew_by_single_year_of_age.json",
                         package = "statxplorerplus")

spec_tbl <- convert_json_to_spec_table(json_path) |>
  add_labels_and_locations_to_spec_table()
```

### Inspect the dates currently in the spec

```{r inspect-dates}
spec_tbl |>
  filter(field_label == "Month") |>
  select(value_label)
```

```
#> # A tibble: 3 × 1
#>   value_label
#>   <chr>
#> 1 August 2013
#> 2 August 2019
#> 3 August 2022
```

The spec only covers three months. Many other months available in the
database are not included.

## Step 3 – Remove the most recent date

To demonstrate what happens when new data becomes available, remove the most
recent date from the spec. This simulates a spec that was saved before the
latest release:

```{r trim}
spec_tbl_trimmed <- spec_tbl |>
  filter(!(field_label == "Month" & value_label == "August 2022"))

spec_tbl_trimmed |>
  filter(field_label == "Month") |>
  select(value_label)
```

```
#> # A tibble: 2 × 1
#>   value_label
#>   <chr>
#> 1 August 2013
#> 2 August 2019
```

## Step 4 – Update to include newer dates

`update_spec_table_to_latest_data()` queries the Stat-Xplore schema for
the date field and appends any date values that are newer than the latest
date already in the spec (here, newer than August 2019).

```{r update}
spec_tbl_updated <- update_spec_table_to_latest_data(spec_tbl_trimmed)
```

### Inspect the updated dates

```{r inspect-updated}
spec_tbl_updated |>
  filter(field_label == "Month") |>
  select(value_label)
```

```
#> # A tibble: 38 × 1
#>    value_label
#>    <chr>
#>  1 August 2013
#>  2 August 2019
#>  3 September 2019
#>  4 October 2019
#>  5 November 2019
#>  6 December 2019
#>  7 January 2020
#>  8 February 2020
#>  9 March 2020
#> 10 April 2020
#> # ℹ 28 more rows
```

The spec now covers 38 months: the original two plus all 36 months from
September 2019 through August 2022.

## Step 5 – Fetch the updated data

```{r fetch}
acc_updated <- fetch_data_from_spec_table(spec_tbl_updated)

head(acc_updated)
```

```
#> # A tibble: 6 × 5
#>   Month       `Age (bands and single year)` `Employment Indicator` `National - Regional - LA - Wards` `Alternative Claimant Count`
#>   <chr>       <chr>                         <chr>                  <chr>                                                     <dbl>
#> 1 August 2013 16                            Not in employment      England                                                     329
#> 2 August 2013 16                            Not in employment      Wales                                                        22
#> 3 August 2013 16                            In employment          England                                                      11
#> 4 August 2013 16                            In employment          Wales                                                         0
#> 5 August 2013 16                            Not available          England                                                       0
#> 6 August 2013 16                            Not available          Wales                                                         0
```

### Check which months are covered

```{r count-months}
acc_updated |> count(Month)
```

```
#> # A tibble: 38 × 2
#>    Month             n
#>    <chr>         <int>
#>  1 April 2020      306
#>  2 April 2021      306
#>  3 April 2022      306
#>  4 August 2013     306
#>  5 August 2019     306
#>  6 August 2020     306
#>  7 August 2021     306
#>  8 August 2022     306
#>  9 December 2019   306
#> 10 December 2020   306
#> # ℹ 28 more rows
```

The data now covers all August 2013 and August 2019, which were in the
original query, and then each of the newer months of data after August
2019 giving 38 months, each with 306 rows (51 age values ×
3 employment statuses × 2 geographies).

## Saving the updated spec back to JSON

After updating, you can export the revised spec so it can be reused or
shared without needing to re-run the update step:

```{r export}
updated_list <- convert_spec_table_to_list(spec_tbl_updated)

export_json(
  .json        = updated_list,
  .output_path = "path/to/acc_ew_by_single_year_of_age_updated.json"
)
```

## Summary

| Step | Function | Purpose |
|---|---|---|
| 1 | `convert_json_to_spec_table()` | Parse JSON → spec table |
| 2 | `add_labels_and_locations_to_spec_table()` | Add human-readable labels |
| 3 | `filter()` | Remove dates to simulate an outdated spec |
| 4 | `update_spec_table_to_latest_data()` | Append date values newer than those in the spec |
| 5 | `fetch_data_from_spec_table()` | Fetch data including new dates |
| 6 | `convert_spec_table_to_list()` + `export_json()` | *(optional)* save updated spec |
