---
output:
  github_document:
    pandoc_args: ["--markdown-headings=atx"]
vignette: >
  %\VignetteIndexEntry{5. Updating a spec to the latest available data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# 5. Updating a spec to the latest available data

## Overview

Stat-Xplore data is updated regularly. A JSON spec you downloaded months ago
will only contain the dates that were available at that time. Rather than
manually editing the JSON every time new data is released,
`update_spec_table_to_latest_data()` queries the Stat-Xplore schema and
appends date values automatically.

This vignette shows the full workflow:

1. Convert a JSON to a spec table.
2. Inspect which dates are currently in the spec.
3. Update the spec to include all dates newer than those already in the spec.
4. Fetch the updated data.

## Step 1 – Load the package and set your API key

```{r load-pkg}
library(statxplorerplus)
library(dplyr)

load_api_key("path/to/apikey.txt")
```

## Step 2 – Convert the JSON to a spec table and add labels

```{r convert}
json_path <- system.file("extdata", "acc_ew_by_single_year_of_age.json",
                         package = "statxplorerplus")

spec_tbl <- convert_json_to_spec_table(json_path) |>
  add_labels_and_locations_to_spec_table()
```

### Inspect the dates currently in the spec

```{r inspect-dates}
spec_tbl |>
  filter(field_label == "Month") |>
  select(value_label)
```

```
#> # A tibble: 3 × 1
#>   value_label
#>   <chr>
#> 1 August 2013
#> 2 August 2019
#> 3 August 2022
```

The spec only covers three months. Many other months available in the
database are not included.

## Step 3 – Update to include newer dates

`update_spec_table_to_latest_data()` queries the Stat-Xplore schema for
the date field and appends any date values that are newer than the latest
date already in the spec.

```{r update}
spec_tbl_updated <- update_spec_table_to_latest_data(spec_tbl)
```

### Inspect the updated dates

```{r inspect-updated}
spec_tbl_updated |>
  filter(field_label == "Month") |>
  select(value_label)
```

```
#> # A tibble: 3 × 1
#>    value_label
#>    <chr>
#>  1 August 2013
#>  2 August 2019
#>  3 August 2022
```

The spec is unchanged because the Alternative Claimant Count database
currently has data up to August 2022, which is already the latest date
in the spec. In practice, once new months are released on Stat-Xplore,
running this step would automatically append them.

## Step 4 – Fetch the updated data

```{r fetch}
acc_updated <- fetch_data_from_spec_table(spec_tbl_updated)

head(acc_updated)
```

```
#> # A tibble: 6 × 5
#>   Month       `Age (bands and single year)` `Employment Indicator` `National - Regional - LA - Wards` `Alternative Claimant Count`
#>   <chr>       <chr>                         <chr>                  <chr>                                                     <dbl>
#> 1 August 2013 16                            Not in employment      England                                                     329
#> 2 August 2013 16                            Not in employment      Wales                                                        22
#> 3 August 2013 16                            In employment          England                                                      11
#> 4 August 2013 16                            In employment          Wales                                                         0
#> 5 August 2013 16                            Not available          England                                                       0
#> 6 August 2013 16                            Not available          Wales                                                         0
```

## Saving the updated spec back to JSON

After updating, you can export the revised spec so it can be reused or
shared without needing to re-run the update step:

```{r export}
updated_list <- convert_spec_table_to_list(spec_tbl_updated)

export_json(
  .json        = updated_list,
  .output_path = "path/to/acc_ew_by_single_year_of_age_updated.json"
)
```

## Summary

| Step | Function | Purpose |
|---|---|---|
| 1 | `convert_json_to_spec_table()` | Parse JSON → spec table |
| 2 | `add_labels_and_locations_to_spec_table()` | Add human-readable labels |
| 3 | `update_spec_table_to_latest_data()` | Append available date values newer than those in the spec |
| 4 | `fetch_data_from_spec_table()` | Fetch data including new dates |
| 5 | `convert_spec_table_to_list()` + `export_json()` | *(optional)* save updated spec |
