---
output:
  github_document:
    pandoc_args: ["--markdown-headings=atx"]
vignette: >
  %\VignetteIndexEntry{6. Building a query from scratch using the schema}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# 6. Building a query from scratch using the schema

## Overview

Sometimes you want to build a Stat-Xplore query from scratch rather than
starting from an existing JSON file. `statxplorerplus` provides a set of
schema-navigation functions that let you browse the available databases,
fields, and values interactively, then turn what you find into a spec table
and a JSON file you can reuse.

This vignette shows how to:

1. List all available databases.
2. Explore measures and fields within a database.
3. Inspect the values available for a field.
4. Assemble a labels table and convert it to a spec table.
5. Export the spec table as a JSON file.
6. Fetch the data.

## Step 1 – Load the package and set your API key

```{r load-pkg}
library(statxplorerplus)
library(dplyr)
library(tibble)

load_api_key("path/to/apikey.txt")
```

## Step 2 – List available databases

```{r list-dbs}
databases <- list_databases()

databases
```

```
#> # A tibble: 120 × 3
#>   id                                       label                                location
#>   <chr>                                    <chr>                                <chr>
#> 1 str:database:ACC                         Alternative Claimant Count           https://...
#> 2 str:database:ACC_OFFS                    Alternative Claimant Count Off Flows https://...
#> 3 str:database:ACC_ONS                     Alternative Claimant Count On Flows  https://...
#> 4 str:database:AA_In_Payment_New           AA: Cases in Payment - Data from...  https://...
#> ...
```

Pick the database you want to work with and save its `location` — this URL
is the entry point for further schema navigation.

```{r pick-db}
acc_location <- databases |>
  filter(label == "Alternative Claimant Count") |>
  pull(location)
```

## Step 3 – Find the measures and fields

`get_target_type_info_below()` traverses the schema tree from a starting
URL and collects all records matching the requested types. Use it to list
all measures and fields in one call:

```{r get-measures-fields}
measures <- get_target_type_info_below(acc_location, c("MEASURE", "COUNT"))
fields   <- get_target_type_info_below(acc_location, "FIELD")

measures
```

```
#> # A tibble: 1 × 3
#>   id                    label                      location
#>   <chr>                 <chr>                      <chr>
#> 1 str:count:ACC:V_F_ACC Alternative Claimant Count https://...
```

```{r show-fields}
fields |> select(id, label, location)
```

```
#> # A tibble: 9 × 3
#>   id                                     label                                  location
#>   <chr>                                  <chr>                                  <chr>
#> 1 str:field:ACC:F_ACC_DATE_new:DATE_NAME Month                                  https://...
#> 2 str:field:ACC:V_F_ACC:AGE              Age (bands and single year)            https://...
#> 3 str:field:ACC:V_F_ACC:CCSEX            Gender                                 https://...
#> 4 str:field:ACC:V_F_ACC:GROUP            Benefit Group                          https://...
#> 5 str:field:ACC:V_F_ACC:EMP              Employment Indicator                   https://...
#> 6 str:field:ACC:V_F_ACC:WARD_CODE        National - Regional - LA - Wards       https://...
#> 7 str:field:ACC:V_F_ACC:UK_COA           National - Regional - LAs - UK         https://...
#> 8 str:field:ACC:V_F_ACC:PARLC_CODE       Westminster Parliamentary Constituency  https://...
#> 9 str:field:ACC:V_F_ACC:GB_COA           National - Regional - LA - COAs - GB   https://...
```

## Step 4 – Inspect the values for a field

Use `get_next_level_info()` to step one level down from a field and see its
valuesets, then step down again to see the individual values.

```{r get-valuesets}
age_location <- fields |>
  filter(label == "Age (bands and single year)") |>
  pull(location)

# One level down: valuesets
age_valuesets <- get_next_level_info(age_location)

age_valuesets |> select(id, label, location)
```

```
#> # A tibble: 2 × 3
#>   id                                            label    location
#>   <chr>                                         <chr>    <chr>
#> 1 str:valueset:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE Age      https://...
#> 2 str:valueset:ACC:V_F_ACC:AGE:C_ACC_AGE_BAND   Age Band https://...
```

```{r get-values}
# Two levels down: individual values (using the first valueset – single years)
age_values <- get_next_level_info(age_valuesets$location[[1]])

age_values |> select(id, label) |> head(10)
```

```
#> # A tibble: 10 × 2
#>   id                                              label
#>   <chr>                                           <chr>
#> 1 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:16  16
#> 2 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:17  17
#> 3 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:18  18
#> 4 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:19  19
#> 5 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:20  20
#> 6 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:21  21
#> 7 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:22  22
#> 8 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:23  23
#> 9 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:24  24
#> 10 str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:25 25
```

Repeat for any other fields you want to include in your query.

## Step 5 – Assemble a labels table

Once you know which fields and values you want, assemble a tibble with one
row per value. The columns must be named `database_label`, `measure_label`,
`field_label`, `valueset_label`, and either `value_label` or `value_code`.

```{r build-labels-tbl}
my_query <- tribble(
  ~database_label,              ~measure_label,
  ~field_label,                               ~valueset_label, ~value_label,
  # Age field – three individual years
  "Alternative Claimant Count", "Alternative Claimant Count",
  "Age (bands and single year)", "Age",        "16",
  "Alternative Claimant Count", "Alternative Claimant Count",
  "Age (bands and single year)", "Age",        "17",
  "Alternative Claimant Count", "Alternative Claimant Count",
  "Age (bands and single year)", "Age",        "18",
  # Geography field – England only
  "Alternative Claimant Count", "Alternative Claimant Count",
  "National - Regional - LA - Wards", "Country", "England",
  # Date field – two months
  "Alternative Claimant Count", "Alternative Claimant Count",
  "Month",                      "Month",       "August 2022",
  "Alternative Claimant Count", "Alternative Claimant Count",
  "Month",                      "Month",       "August 2023"
)
```

## Step 6 – Convert to a spec table

`convert_labels_table_to_spec_table()` uses fuzzy matching to look up each
label in the schema and return the corresponding Stat-Xplore IDs. It will
warn you if any label matches only approximately, and error if it cannot
find a match.

```{r convert-labels}
spec_tbl <- convert_labels_table_to_spec_table(my_query)

spec_tbl |> select(field_id, value_id, value_label)
```

```
#> # A tibble: 6 × 3
#>   field_id                               value_id                                                           value_label
#>   <chr>                                  <chr>                                                              <chr>
#> 1 str:field:ACC:V_F_ACC:AGE              str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:16                     16
#> 2 str:field:ACC:V_F_ACC:AGE              str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:17                     17
#> 3 str:field:ACC:V_F_ACC:AGE              str:value:ACC:V_F_ACC:AGE:C_ACC_SINGLE_AGE:18                     18
#> 4 str:field:ACC:V_F_ACC:WARD_CODE        str:value:ACC:V_F_ACC:WARD_CODE:V_C_MASTERGEOG11_COUNTRY_TO_UK:E92000001  England
#> 5 str:field:ACC:F_ACC_DATE_new:DATE_NAME str:value:ACC:F_ACC_DATE_new:DATE_NAME:C_ACC_DATE:202208           August 2022
#> 6 str:field:ACC:F_ACC_DATE_new:DATE_NAME str:value:ACC:F_ACC_DATE_new:DATE_NAME:C_ACC_DATE:202308           August 2023
```

## Step 7 – Export as JSON

Convert the spec table to a list and write it to a JSON file that can be
reused later (e.g. with `fetch_table(filename = ...)` or shared with
colleagues).

```{r export}
json_list <- convert_spec_table_to_list(spec_tbl)

export_json(
  .json        = json_list,
  .output_path = "path/to/my_new_query.json"
)
```

## Step 8 – Fetch the data

Fetch directly from the spec table without the round-trip through JSON:

```{r fetch}
my_data <- fetch_data_from_spec_table(spec_tbl)

head(my_data)
```

```
#>   measure_label              AGE DATE_NAME    WARD_CODE value
#>   <chr>                      <chr> <chr>      <chr>     <dbl>
#> 1 Alternative Claimant Count 16   August 2022 England    1042
#> 2 Alternative Claimant Count 16   August 2023 England    1089
#> ...
```

## Summary

| Step | Function | Purpose |
|---|---|---|
| 1 | `list_databases()` | Browse all available databases |
| 2 | `get_target_type_info_below()` | List all measures and fields |
| 3 | `get_next_level_info()` | Inspect valuesets and values |
| 4 | `tribble()` / `tibble()` | Assemble a labels table |
| 5 | `convert_labels_table_to_spec_table()` | Resolve labels to IDs |
| 6 | `convert_spec_table_to_list()` + `export_json()` | Save as JSON |
| 7 | `fetch_data_from_spec_table()` | Fetch data from API |
